---
title: "Bayesian and bootstrap probability estimates with fitPS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian and bootstrap probability estimates with fitPS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
suppressPackageStartupMessages(library(fitPS))
```

## Roux et al. footwear survey

We illustrate the functionality using the Roux et al. (2001) footwear data set, supplied with `fitPS` as a P survey. In this survey, the response is the number of different sources of glass found on each surveyed pair of shoes.

```{r roux-data}
data("Psurveys")
roux = Psurveys$roux
roux
```

The zero-inflated zeta model has two parameters: the inflation probability `pi` and the standard fitPS zeta `shape` parameter, which is greater than 1.

## Plug-in probabilities

The conventional maximum-likelihood fit evaluates each P term at the fitted parameter values.

```{r mle-fit}
mleFit = fitZIDist(roux, nterms = 6)
fitted(mleFit, type = "plugIn")
```

These are plug-in estimates of the form

\[
P_k(\widehat{\theta}),
\]

where \(\widehat{\theta}=(\widehat{\pi},\widehat{\mathrm{shape}})\).

## Bayesian posterior probabilities

A Bayesian zero-inflated fit is requested with `method = "bayes"`. The default posterior engine is deterministic numerical integration over the joint posterior.

```{r bayes-fit}
bayesFit = fitZIDist(
  roux,
  nterms = 6,
  method = "bayes",
  bayesOptions = list(
    posteriorMethod = "numerical"
  )
)

bayesFit$posterior
```

The default prior for `pi` is Beta(1, 1). The shape prior is created by `makePrior()` and is log-uniform on `shape - 1` over its default support. Alternative beta-prior parameters can be supplied with `shape1` and `shape2`, and an alternative shape prior can be supplied through `bayesOptions$prior`.

The probability summaries are obtained directly from the joint posterior distribution of `pi` and `shape`.

```{r posterior-probs}
posteriorProbs(bayesFit, n = 6)
```

The `estimate` column is the posterior mean probability,

\[
E\{P_k(\theta)\mid x\},
\]

not the probability evaluated at posterior mean parameters. Because the P terms are nonlinear functions of the parameters, those two quantities need not agree.

```{r posterior-vs-plugin}
cbind(
  plugIn = fitted(bayesFit, n = 6, type = "plugIn"),
  posteriorMean = fitted(bayesFit, n = 6, type = "posteriorMean")
)
```

The posterior object can also be plotted directly. Points are posterior means and vertical lines are equal-tailed credible intervals.

```{r posterior-plot}
plot(bayesFit$posterior, n = 6)
```

### Inference about the inflation component

The inflation parameter `pi` is itself scientifically useful. It measures the extra probability assigned to the inflated value beyond that supplied by the zeta component. For a P survey this is extra mass at `P0`; for an S survey it is extra mass at `S1`. The marginal posterior for `pi` can therefore be inspected directly.

```{r pi-posterior}
plotPosterior(bayesFit, parameter = "pi")
```

If the posterior for `pi` is concentrated near zero, with little support for appreciable inflation, that is evidence that the additional zero/one-inflation component may not be needed in practice. Conversely, posterior mass away from zero supports a substantive inflation component. This should be treated as a diagnostic rather than a formal posterior probability that the simpler model is true: under the continuous beta prior used here, the posterior probability of exactly `pi = 0` is zero. A formal Bayesian probability for the no-inflation model would require an explicit point mass at zero, a spike-and-slab construction, or comparison of the ordinary and inflated models.

### Other posterior engines

The same public posterior interface is used for all four Bayesian engines. The numerical method is the default and is used above because it is deterministic. The alternatives are:

```{r other-engines, eval=FALSE}
mcmcFit = fitZIDist(
  roux,
  method = "bayes",
  bayesOptions = list(posteriorMethod = "mcmc"),
  nIter = 5000,
  nBurnIn = 1000,
  seed = 1234
)

laplaceFit = fitZIDist(
  roux,
  method = "bayes",
  bayesOptions = list(posteriorMethod = "laplace"),
  seed = 1234
)

importanceFit = fitZIDist(
  roux,
  method = "bayes",
  bayesOptions = list(posteriorMethod = "importance"),
  nSamples = 5000,
  seed = 1234
)
```

For the numerical engine, P-term expectations are integrated over the weighted joint posterior grid. For MCMC, each retained `(pi, shape)` draw is transformed into P terms. Importance sampling retains and uses the importance weights in means, variances, and quantiles. The Laplace engine simulates seeded draws on its unconstrained Gaussian approximation and transforms those draws back to `(pi, shape)` before calculating P terms.

## Bootstrap probability distributions

The frequentist analogue uses the nonparametric bootstrap distribution of the fitted parameter estimator. Starting from the maximum-likelihood fit, `bootstrapFit()` repeatedly resamples the Roux survey observations, refits the model, and transforms each successful bootstrap parameter replicate into P terms.

```{r bootstrap-fit}
bootFit = bootstrapFit(
  mleFit,
  B = 100,
  seed = 1234,
  silent = TRUE,
  parallel = FALSE
)

bootstrapProbs(bootFit, n = 6)
```

The bootstrap probability estimate is

\[
E_*\{P_k(\widehat{\theta}^*)\},
\]

where \(E_*\) denotes expectation over the bootstrap distribution. It is not a Bayesian posterior mean and it is generally not identical to the plug-in probability.

There is nevertheless a useful Bayesian perspective on the bootstrap. Efron described the bootstrap as moving from the observed-data world to a "bootstrap world" by replacing the unknown data-generating distribution with an estimated one, and later showed that parametric-bootstrap draws can be reweighted by importance sampling to compute Bayesian posterior quantities. Thus bootstrap and posterior simulation remain different inferential procedures, but they support closely parallel calculations for derived quantities. In `fitPS`, the frequentist calculation averages \(P_k(\widehat{\theta}^*)\) over bootstrap replicates, whereas the Bayesian calculation averages \(P_k(\theta)\) over the posterior distribution. This parallel is one reason the package represents bootstrap and posterior uncertainty using closely related objects while retaining their different interpretations.

```{r compare-three}
cbind(
  plugIn = fitted(mleFit, n = 6, type = "plugIn"),
  bootstrapMean = fitted(bootFit, n = 6, type = "bootstrapMean"),
  posteriorMean = fitted(bayesFit, n = 6, type = "posteriorMean")
)
```

The bootstrap probability object reports percentile confidence intervals and can be plotted in the same style as the posterior object.

```{r bootstrap-plot}
plot(bootFit$bootstrap, n = 6)
```

For applied work, substantially more than 100 bootstrap replicates would normally be used. The smaller value here keeps vignette building practical while demonstrating the complete workflow on the real Roux data.

## Prediction interfaces

The same distinction is available through `predict()`.

```{r predict-interface}
predict(mleFit, newdata = 0:5, type = "plugIn")
predict(bootFit, newdata = 0:5, type = "bootstrapMean", interval = "percentile")
predict(bayesFit, newdata = 0:5, type = "posteriorMean", interval = "credible")
```

Thus the three probability summaries have distinct interpretations:

\[
P_k(\widehat{\theta}), \qquad
E_*\{P_k(\widehat{\theta}^*)\}, \qquad
E\{P_k(\theta)\mid x\}.
\]

`probfun()` remains the plug-in probability interface. Use `posteriorProbs()` for Bayesian posterior probability summaries and `bootstrapProbs()` for bootstrap probability summaries.

## Reference

Efron, B. (2003). Second thoughts on the bootstrap. *Statistical Science*, 18(2), 135-140.

Efron, B. (2012). Bayesian inference and the parametric bootstrap. *The Annals of Applied Statistics*, 6(4), 1971-1997.

Roux, C., Kirk, R., Benson, S., Van Haren, T., and Petterd, C. I. (2001). Glass particles in footwear of members of the public in south-eastern Australia: a survey. *Forensic Science International*, 116(2), 149-156.
