---
title: "Extending fitPS with External Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending fitPS with External Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
suppressPackageStartupMessages(library(fitPS))
```

## Why this vignette exists

`fitPS` supports models defined outside the package. A downstream package or an ordinary R script can construct a `psModel` subclass, provide methods for a small public S3 contract, and pass that model directly to `fit()`.

The important architectural point is that the distribution-specific mathematics stays outside `fitPS`. The package supplies the generic fitting and fitted-object machinery; the external model supplies its parameter names, observation mapping, probabilities, likelihood, and any model-specific maximum-likelihood controls it needs.

This vignette develops two examples:

1. an ordinary Poisson model, which demonstrates the smallest useful one-parameter extension; and
2. a two-parameter Poisson-normal (Hermite) model parameterized by `mu` and `sigma`, which demonstrates that the same public API can handle a more demanding model with a coupled parameter constraint.

Neither example requires editing or rebuilding `fitPS`.

## The public model contract

An external maximum-likelihood model normally needs the following public pieces:

- `psModel()` to construct the model descriptor;
- `modelObservationData()` to map a `psData` object to the values used by the distribution;
- `modelProbabilities()` to calculate fitted or predicted probability terms;
- `modelLogLikelihood()` to calculate the model log likelihood; and
- optionally `modelMleControl()` when the defaults stored in the model descriptor are not sufficient.

The descriptor also records parameter names. `supportedPosteriorEngines()` reports any Bayesian engines explicitly supported by the model. An external model does not need to advertise a Bayesian engine simply to participate in maximum-likelihood fitting.

Once these pieces are supplied, `fit()` returns an ordinary `psFit`. Generic operations such as `fitted()`, `predict()`, `logLik()`, `deviance()`, `AIC()`, and `BIC()` continue to be handled by `fitPS`.

## Example 1: an external Poisson model

### Construct the model

The model constructor contains only declarative information. The Poisson model has one parameter, `lambda`, and uses a positive lower bound for generic maximum-likelihood optimisation.

```{r poisson-constructor}
externalPoissonModel = function() {
  psModel(
    model = "poisson",
    parameterNames = "lambda",
    subclass = "externalPoissonModel",
    mleStart = c(lambda = 1),
    mleLower = c(lambda = sqrt(.Machine$double.eps))
  )
}
```

The resulting object is a `psModel` subclass, but it is not a built-in `fitPS` distribution.

```{r poisson-descriptor}
poissonModel = externalPoissonModel()
class(poissonModel)
modelParameterNames(poissonModel)
supportedPosteriorEngines(poissonModel)
```

An empty value from `supportedPosteriorEngines()` is intentional here: this example defines an external MLE model only.

### Supply the observation mapping

For these P-survey data, the observed count value used by the Poisson mass function is stored in `x$data$n`.

```{r poisson-observations}
modelObservationData.externalPoissonModel = function(model, x, ...) {
  x$data$n
}
```

### Supply probabilities

`modelProbabilities()` receives named parameter values and the probability terms requested by the generic fitted/prediction machinery.

```{r poisson-probabilities}
modelProbabilities.externalPoissonModel = function(model,
                                                     parameters,
                                                     n,
                                                     type,
                                                     ...) {
  lambda = parameters[["lambda"]]
  values = vapply(
    n,
    function(value) {
      dpois(value, lambda = lambda)
    },
    numeric(length(lambda))
  )

  if (length(lambda) == 1L) {
    values = matrix(values, nrow = 1L)
  }

  colnames(values) = paste0(type, n)
  values
}
```

### Supply the log likelihood

The likelihood is weighted by the survey frequencies stored in the `psData` object.

```{r poisson-likelihood}
modelLogLikelihood.externalPoissonModel = function(model,
                                                    parameters,
                                                    data,
                                                    ...) {
  lambda = parameters[["lambda"]]
  observations = modelObservationData(model, data)
  sum(data$data$rn * dpois(observations, lambda = lambda, log = TRUE))
}
```

### Register the external S3 methods

A downstream package should register these methods in its `NAMESPACE`, for example:

```text
S3method(modelObservationData,externalPoissonModel)
S3method(modelProbabilities,externalPoissonModel)
S3method(modelLogLikelihood,externalPoissonModel)
```

When experimenting interactively, the equivalent registration can be performed with base R's `registerS3method()`:

```{r poisson-register}
registerS3method(
  "modelObservationData",
  "externalPoissonModel",
  modelObservationData.externalPoissonModel,
  envir = asNamespace("fitPS")
)
registerS3method(
  "modelProbabilities",
  "externalPoissonModel",
  modelProbabilities.externalPoissonModel,
  envir = asNamespace("fitPS")
)
registerS3method(
  "modelLogLikelihood",
  "externalPoissonModel",
  modelLogLikelihood.externalPoissonModel,
  envir = asNamespace("fitPS")
)
```

No `fitPS:::` access is required.

### Fit and use the model

Create a small P-survey data set and pass the external model directly to `fit()`.

```{r poisson-fit}
poissonData = makePSData(
  n = c(0, 1, 2, 3),
  count = c(30, 12, 5, 1),
  type = "P"
)

poissonFit = fit(
  poissonData,
  model = externalPoissonModel(),
  nterms = 5
)

poissonFit$lambda
```

The result is an ordinary `psFit`, and it retains the originating external model object.

```{r poisson-fit-contract}
class(poissonFit)
class(poissonFit$modelObject)
fitted(poissonFit)
predict(poissonFit, newdata = 0:4, interval = "none")
```

The common model-comparison infrastructure also works without any Poisson-specific AIC or BIC implementation.

```{r poisson-comparison}
logLik(poissonFit)
deviance(poissonFit)
AIC(poissonFit)
BIC(poissonFit)
```

## Example 2: a Poisson-normal model with `mu` and `sigma`

The second example deliberately places more pressure on the public contract. It has two parameters and a valid region in which the two parameters are coupled.

Here the Poisson-normal model is the Hermite count distribution represented as

\[
X = X_1 + 2X_2,
\]

where `X1` and `X2` are independent Poisson random variables. We expose the distribution through the more familiar parameters

\[
\mu = E(X), \qquad \sigma = \sqrt{\operatorname{Var}(X)}.
\]

For this representation,

\[
\lambda_2 = \frac{\sigma^2 - \mu}{2},
\qquad
\lambda_1 = 2\mu - \sigma^2.
\]

Both component rates must be non-negative, so the valid parameter region is

\[
\mu \leq \sigma^2 \leq 2\mu.
\]

The public parameter names are therefore `mu` and `sigma`, analogous to the naming used by `dnorm()`.

### Construct the two-parameter descriptor

```{r poisson-normal-constructor}
externalPoissonNormalModel = function() {
  psModel(
    model = "poissonNormal",
    parameterNames = c("mu", "sigma"),
    subclass = "externalPoissonNormalModel"
  )
}
```

### Map the data and choose starting values

The same observation mapping is appropriate. This model also supplies `modelMleControl()` because useful starting values depend on the observed data.

```{r poisson-normal-control}
modelObservationData.externalPoissonNormalModel = function(model, x, ...) {
  x$data$n
}

modelMleControl.externalPoissonNormalModel = function(model, x, ...) {
  observations = modelObservationData(model, x)
  weights = x$data$rn
  muStart = weighted.mean(observations, weights)
  varianceStart = weighted.mean((observations - muStart)^2, weights)
  lowerVariance = muStart * 1.05
  upperVariance = muStart * 1.95
  varianceStart = min(max(varianceStart, lowerVariance), upperVariance)

  list(
    start = c(mu = muStart, sigma = sqrt(varianceStart)),
    lower = c(
      mu = sqrt(.Machine$double.eps),
      sigma = sqrt(.Machine$double.eps)
    ),
    upper = c(mu = Inf, sigma = Inf)
  )
}
```

### Implement the distribution mathematics externally

The helper below converts `mu` and `sigma` to the two Poisson rates. Returning `NULL` outside the legal region lets the likelihood reject invalid trial points without requiring a Poisson-normal special case inside `fitPS`.

```{r poisson-normal-rates}
poissonNormalComponentRates = function(mu, sigma) {
  variance = sigma^2
  lambdaTwo = (variance - mu) / 2
  lambdaOne = 2 * mu - variance

  if (!is.finite(mu) || !is.finite(sigma) || mu <= 0 || sigma <= 0 ||
      lambdaOne < 0 || lambdaTwo < 0) {
    return(NULL)
  }

  c(lambdaOne = lambdaOne, lambdaTwo = lambdaTwo)
}
```

The probability mass at an integer `n` is obtained by summing over all possible values of `X2`.

```{r poisson-normal-pmf}
poissonNormalProbability = function(n, mu, sigma) {
  rates = poissonNormalComponentRates(mu, sigma)

  if (is.null(rates)) {
    return(rep(NaN, length(n)))
  }

  vapply(n, function(value) {
    if (!is.finite(value) || value < 0 || value != floor(value)) {
      return(0)
    }

    doubleCount = 0:floor(value / 2)
    singleCount = value - 2 * doubleCount

    sum(
      dpois(singleCount, lambda = rates[["lambdaOne"]]) *
        dpois(doubleCount, lambda = rates[["lambdaTwo"]])
    )
  }, numeric(1L))
}
```

The two model methods simply expose that external mathematics to the fitPS generics.

```{r poisson-normal-methods}
modelProbabilities.externalPoissonNormalModel = function(model,
                                                          parameters,
                                                          n,
                                                          type,
                                                          ...) {
  mu = parameters[["mu"]]
  sigma = parameters[["sigma"]]
  values = poissonNormalProbability(n, mu = mu, sigma = sigma)
  values = matrix(values, nrow = 1L)
  colnames(values) = paste0(type, n)
  values
}

modelLogLikelihood.externalPoissonNormalModel = function(model,
                                                           parameters,
                                                           data,
                                                           ...) {
  mu = parameters[["mu"]]
  sigma = parameters[["sigma"]]
  observations = modelObservationData(model, data)
  probabilities = poissonNormalProbability(
    observations,
    mu = mu,
    sigma = sigma
  )

  if (any(!is.finite(probabilities)) || any(probabilities <= 0)) {
    return(-Inf)
  }

  sum(data$data$rn * log(probabilities))
}
```

### Register the Poisson-normal methods

A downstream package would normally use these namespace registrations:

```text
S3method(modelObservationData,externalPoissonNormalModel)
S3method(modelMleControl,externalPoissonNormalModel)
S3method(modelProbabilities,externalPoissonNormalModel)
S3method(modelLogLikelihood,externalPoissonNormalModel)
```

For this self-contained vignette we register them directly:

```{r poisson-normal-register}
registerS3method(
  "modelObservationData",
  "externalPoissonNormalModel",
  modelObservationData.externalPoissonNormalModel,
  envir = asNamespace("fitPS")
)
registerS3method(
  "modelMleControl",
  "externalPoissonNormalModel",
  modelMleControl.externalPoissonNormalModel,
  envir = asNamespace("fitPS")
)
registerS3method(
  "modelProbabilities",
  "externalPoissonNormalModel",
  modelProbabilities.externalPoissonNormalModel,
  envir = asNamespace("fitPS")
)
registerS3method(
  "modelLogLikelihood",
  "externalPoissonNormalModel",
  modelLogLikelihood.externalPoissonNormalModel,
  envir = asNamespace("fitPS")
)
```

### Fit the two-parameter model

```{r poisson-normal-fit}
poissonNormalData = makePSData(
  n = 0:9,
  count = c(223, 223, 223, 149, 93, 48, 24, 10, 4, 2),
  type = "P"
)

poissonNormalFit = fit(
  poissonNormalData,
  model = externalPoissonNormalModel(),
  nterms = 10
)

c(mu = poissonNormalFit$mu, sigma = poissonNormalFit$sigma)
```

The fitted values should satisfy the model's legal region.

```{r poisson-normal-region}
poissonNormalFit$mu <= poissonNormalFit$sigma^2
poissonNormalFit$sigma^2 <= 2 * poissonNormalFit$mu
```

The same common fitted-object and model-comparison APIs remain available.

```{r poisson-normal-generic-apis}
fitted(poissonNormalFit)
predict(poissonNormalFit, newdata = 0:9, interval = "none")
logLik(poissonNormalFit)
deviance(poissonNormalFit)
AIC(poissonNormalFit)
BIC(poissonNormalFit)
```

## Serialization and retained model objects

New `psFit` objects retain their originating model descriptor. That avoids reconstructing third-party models from a hard-coded list of model names.

```{r retained-model}
class(poissonFit$modelObject)
class(poissonNormalFit$modelObject)
```

This also means a fitted object can survive ordinary R serialization, provided the corresponding S3 methods are available when the object is used again.

```{r serialization}
path = tempfile(fileext = ".rds")
saveRDS(poissonNormalFit, path)
restoredFit = readRDS(path)
logLik(restoredFit)
```

A downstream package therefore needs to remain installed and loadable when its fitted objects are restored, just as with other S3-based extension systems.

## What an extension package should contain

A real external package does not need to copy fitPS fitting code. Its distribution layer can remain small:

1. a constructor returning `psModel(...)` with a unique subclass;
2. S3 methods for the required public fitPS generics;
3. any private helpers used to evaluate probabilities or parameter transformations;
4. `S3method(...)` entries in its `NAMESPACE`; and
5. behavioural tests showing that `fit()`, fitted probabilities, prediction, and relevant model-comparison criteria work.

For an MLE-only model, there is no requirement to implement a posterior engine. Bayesian support should be declared only when the model supplies the corresponding engine-specific contract.

## Design lessons from the two examples

The Poisson example shows that a simple distribution can be added without another public fitting function and without any fitPS source modification.

The Poisson-normal example demonstrates something stronger. Its distribution mathematics, parameter transformation, starting-value logic, and legal parameter region all remain external, while fitPS still provides the optimizer orchestration and the standard `psFit` interface. No Poisson-normal branch is needed in `fit()`, `predict()`, `logLik()`, AIC, BIC, or deviance.

That is the intended extension boundary: **new statistical models supply model mathematics; fitPS supplies common fitting and fitted-object infrastructure.**
