---
title: "Extending fitPS with External Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending fitPS with External Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
suppressPackageStartupMessages(library(fitPS))
```

## Why this vignette exists

`fitPS` supports models defined outside the package. A downstream package or an ordinary R script can construct a `psModel` subclass, provide methods for a small public S3 contract, and pass that model directly to `fit()`.

The important architectural point is that the distribution-specific mathematics stays outside `fitPS`. The package supplies the generic fitting and fitted-object machinery; the external model supplies its parameter names, observation mapping, probabilities, likelihood, and any model-specific maximum-likelihood controls it needs.

This vignette develops two examples:

1. an ordinary Poisson model, which demonstrates the smallest useful one-parameter extension; and
2. a two-parameter Poisson-normal model parameterized by `mu` and `sigma`, in which a latent normal log-rate is integrated out of the Poisson likelihood.

Neither example requires editing or rebuilding `fitPS`.

## The public model contract

An external maximum-likelihood model normally needs the following public pieces:

- `psModel()` to construct the model descriptor;
- `modelObservationData()` to map a `psData` object to the values used by the distribution;
- `modelProbabilities()` to calculate fitted or predicted probability terms;
- `modelLogLikelihood()` to calculate the model log likelihood; and
- optionally `modelMleControl()` when the defaults stored in the model descriptor are not sufficient.

The descriptor also records parameter names. `supportedPosteriorEngines()` reports any Bayesian engines explicitly supported by the model. An external model does not need to advertise a Bayesian engine simply to participate in maximum-likelihood fitting.

Once these pieces are supplied, `fit()` returns an ordinary `psFit`. Generic operations such as `fitted()`, `predict()`, `logLik()`, `deviance()`, `AIC()`, and `BIC()` continue to be handled by `fitPS`.

## Example 1: an external Poisson model

For a Poisson random variable with rate `lambda`,

\[
P(Y = y \mid \lambda) = \frac{e^{-\lambda}\lambda^y}{y!},
\qquad y = 0, 1, 2, \ldots
\]

If the observed count value `y_i` occurs with survey frequency `r_i`, the weighted log likelihood used by these examples is

\[
\ell(\lambda) = \sum_i r_i
\left[
-\lambda + y_i \log(\lambda) - \log(y_i!)
\right].
\]

This closed-form probability mass function makes Poisson a useful minimal example of the public extension contract.

### Construct the model

The model constructor contains only declarative information. The Poisson model has one parameter, `lambda`, and uses a positive lower bound for generic maximum-likelihood optimisation.

```{r poisson-constructor}
externalPoissonModel = function() {
  psModel(
    model = "poisson",
    parameterNames = "lambda",
    subclass = "externalPoissonModel",
    mleStart = c(lambda = 1),
    mleLower = c(lambda = sqrt(.Machine$double.eps))
  )
}
```

The resulting object is a `psModel` subclass, but it is not a built-in `fitPS` distribution.

```{r poisson-descriptor}
poissonModel = externalPoissonModel()
class(poissonModel)
modelParameterNames(poissonModel)
supportedPosteriorEngines(poissonModel)
```

An empty value from `supportedPosteriorEngines()` is intentional here: this example defines an external MLE model only.

### Supply the observation mapping

For these P-survey data, the observed count value used by the Poisson mass function is stored in `x$data$n`.

```{r poisson-observations}
modelObservationData.externalPoissonModel = function(model, x, ...) {
  x$data$n
}
```

### Supply probabilities

`modelProbabilities()` receives named parameter values and the probability terms requested by the generic fitted/prediction machinery.

```{r poisson-probabilities}
modelProbabilities.externalPoissonModel = function(model,
                                                     parameters,
                                                     n,
                                                     type,
                                                     ...) {
  lambda = parameters[["lambda"]]
  values = vapply(
    n,
    function(value) {
      dpois(value, lambda = lambda)
    },
    numeric(length(lambda))
  )

  if (length(lambda) == 1L) {
    values = matrix(values, nrow = 1L)
  }

  colnames(values) = paste0(type, n)
  values
}
```

### Supply the log likelihood

The likelihood is weighted by the survey frequencies stored in the `psData` object.

```{r poisson-likelihood}
modelLogLikelihood.externalPoissonModel = function(model,
                                                    parameters,
                                                    data,
                                                    ...) {
  lambda = parameters[["lambda"]]
  observations = modelObservationData(model, data)
  sum(data$data$rn * dpois(observations, lambda = lambda, log = TRUE))
}
```

### Register the external S3 methods

A downstream package should register these methods in its `NAMESPACE`, for example:

```text
S3method(modelObservationData,externalPoissonModel)
S3method(modelProbabilities,externalPoissonModel)
S3method(modelLogLikelihood,externalPoissonModel)
```

When experimenting interactively, the equivalent registration can be performed with base R's `registerS3method()`:

```{r poisson-register}
registerS3method(
  "modelObservationData",
  "externalPoissonModel",
  modelObservationData.externalPoissonModel,
  envir = asNamespace("fitPS")
)
registerS3method(
  "modelProbabilities",
  "externalPoissonModel",
  modelProbabilities.externalPoissonModel,
  envir = asNamespace("fitPS")
)
registerS3method(
  "modelLogLikelihood",
  "externalPoissonModel",
  modelLogLikelihood.externalPoissonModel,
  envir = asNamespace("fitPS")
)
```

No `fitPS:::` access is required.

### Fit and use the model

Create a small P-survey data set and pass the external model directly to `fit()`.

```{r poisson-fit}
poissonData = makePSData(
  n = c(0, 1, 2, 3),
  count = c(30, 12, 5, 1),
  type = "P"
)

poissonFit = fit(
  poissonData,
  model = externalPoissonModel(),
  nterms = 5
)

poissonFit$lambda
```

The result is an ordinary `psFit`, and it retains the originating external model object.

```{r poisson-fit-contract}
class(poissonFit)
class(poissonFit$modelObject)
fitted(poissonFit)
predict(poissonFit, newdata = 0:4, interval = "none")
```

The common model-comparison infrastructure also works without any Poisson-specific AIC or BIC implementation.

```{r poisson-comparison}
logLik(poissonFit)
deviance(poissonFit)
AIC(poissonFit)
BIC(poissonFit)
```

## Example 2: a Poisson-normal model with `mu` and `sigma`

The second example is deliberately more demanding because its marginal likelihood contains an integral.

Let the latent log-rate be normal,

\[
Z \sim N(\mu, \sigma^2),
\]

and let the observed count be conditionally Poisson,

\[
Y \mid Z = z \sim \operatorname{Poisson}(e^z).
\]

The parameters `mu` and `sigma` therefore have the same interpretation and naming convention as the corresponding arguments to `dnorm()`: they describe the mean and standard deviation of the latent normal variable `Z`.

After integrating out `Z`, the marginal probability of a count `y` is

\[
P(Y = y \mid \mu, \sigma)
=
\int_{-\infty}^{\infty}
\frac{\exp(-e^z)e^{yz}}{y!}
\frac{1}{\sigma\sqrt{2\pi}}
\exp\left\{-\frac{(z-\mu)^2}{2\sigma^2}\right\}
\, dz.
\]

Equivalently,

\[
P(Y = y \mid \mu, \sigma)
=
\int_{-\infty}^{\infty}
P(Y=y\mid Z=z) f_Z(z;\mu,\sigma)\,dz.
\]

There is no simple closed-form expression for this integral, so this example evaluates it numerically. If `y_i` occurs with frequency `r_i`, the marginal log likelihood is

\[
\ell(\mu,\sigma)
=
\sum_i r_i
\log\left\{
\int_{-\infty}^{\infty}
P(Y=y_i\mid Z=z) f_Z(z;\mu,\sigma)\,dz
\right\},
\qquad \sigma > 0.
\]

This is a substantially stronger extensibility test than the ordinary Poisson model: the numerical integration remains entirely in downstream model code while `fitPS` continues to own the generic optimizer and fitted-object machinery.

### Construct the two-parameter descriptor

```{r poisson-normal-constructor}
externalPoissonNormalModel = function() {
  psModel(
    model = "poissonNormal",
    parameterNames = c("mu", "sigma"),
    subclass = "externalPoissonNormalModel"
  )
}
```

### Map the data and choose starting values

The observation mapping is unchanged. Starting values can be obtained from the marginal mean and variance of the Poisson-normal model,

\[
E(Y) = \exp\left(\mu + \frac{\sigma^2}{2}\right),
\]

and

\[
\operatorname{Var}(Y)
=
E(Y)
+
E(Y)^2\left\{\exp(\sigma^2)-1\right\}.
\]

These moment relationships provide convenient initial values for `mu` and `sigma` without changing the model parameterization.

```{r poisson-normal-control}
modelObservationData.externalPoissonNormalModel = function(model, x, ...) {
  x$data$n
}

modelMleControl.externalPoissonNormalModel = function(model, x, ...) {
  observations = modelObservationData(model, x)
  weights = x$data$rn
  meanStart = weighted.mean(observations, weights)
  varianceStart = weighted.mean((observations - meanStart)^2, weights)

  if (!is.finite(meanStart) || meanStart <= 0) {
    meanStart = 1
  }
  if (!is.finite(varianceStart)) {
    varianceStart = meanStart
  }

  extraVariance = max(
    varianceStart - meanStart,
    meanStart^2 * 1e-6
  )
  sigmaSquaredStart = log1p(extraVariance / meanStart^2)
  sigmaStart = sqrt(max(sigmaSquaredStart, 1e-6))
  muStart = log(meanStart) - sigmaSquaredStart / 2

  list(
    start = c(mu = muStart, sigma = sigmaStart),
    lower = c(mu = -20, sigma = sqrt(.Machine$double.eps)),
    upper = c(mu = 20, sigma = 5)
  )
}
```

### Evaluate the marginal probability by numerical integration

The distribution-specific numerical integration is downstream code. `fitPS` does not need a Poisson-normal branch or a special integration engine for this model.

```{r poisson-normal-pmf}
poissonNormalProbability = function(n, mu, sigma) {
  if (!is.finite(mu) || !is.finite(sigma) || sigma <= 0) {
    return(rep(NaN, length(n)))
  }

  vapply(n, function(value) {
    if (!is.finite(value) || value < 0 || value != floor(value)) {
      return(0)
    }

    integrand = function(z) {
      dpois(value, lambda = exp(z)) *
        dnorm(z, mean = mu, sd = sigma)
    }

    result = integrate(
      integrand,
      lower = -Inf,
      upper = Inf,
      rel.tol = 1e-8,
      subdivisions = 200L,
      stop.on.error = FALSE
    )

    if (!identical(result$message, "OK")) {
      return(NaN)
    }

    result$value
  }, numeric(1L))
}
```

The fitPS-facing methods simply expose those marginal probabilities and their weighted log likelihood.

```{r poisson-normal-methods}
modelProbabilities.externalPoissonNormalModel = function(model,
                                                          parameters,
                                                          n,
                                                          type,
                                                          ...) {
  mu = parameters[["mu"]]
  sigma = parameters[["sigma"]]
  values = poissonNormalProbability(n, mu = mu, sigma = sigma)
  values = matrix(values, nrow = 1L)
  colnames(values) = paste0(type, n)
  values
}

modelLogLikelihood.externalPoissonNormalModel = function(model,
                                                           parameters,
                                                           data,
                                                           ...) {
  mu = parameters[["mu"]]
  sigma = parameters[["sigma"]]
  observations = modelObservationData(model, data)
  probabilities = poissonNormalProbability(
    observations,
    mu = mu,
    sigma = sigma
  )

  if (any(!is.finite(probabilities)) || any(probabilities <= 0)) {
    return(-Inf)
  }

  sum(data$data$rn * log(probabilities))
}
```

### Register the Poisson-normal methods

A downstream package would normally use these namespace registrations:

```text
S3method(modelObservationData,externalPoissonNormalModel)
S3method(modelMleControl,externalPoissonNormalModel)
S3method(modelProbabilities,externalPoissonNormalModel)
S3method(modelLogLikelihood,externalPoissonNormalModel)
```

For this self-contained vignette we register them directly:

```{r poisson-normal-register}
registerS3method(
  "modelObservationData",
  "externalPoissonNormalModel",
  modelObservationData.externalPoissonNormalModel,
  envir = asNamespace("fitPS")
)
registerS3method(
  "modelMleControl",
  "externalPoissonNormalModel",
  modelMleControl.externalPoissonNormalModel,
  envir = asNamespace("fitPS")
)
registerS3method(
  "modelProbabilities",
  "externalPoissonNormalModel",
  modelProbabilities.externalPoissonNormalModel,
  envir = asNamespace("fitPS")
)
registerS3method(
  "modelLogLikelihood",
  "externalPoissonNormalModel",
  modelLogLikelihood.externalPoissonNormalModel,
  envir = asNamespace("fitPS")
)
```

### Fit the two-parameter model

The following frequencies were generated from a Poisson-normal distribution with values close to `mu = 0.2` and `sigma = 0.45`.

```{r poisson-normal-fit}
poissonNormalData = makePSData(
  n = 0:8,
  count = c(3032, 3240, 2035, 997, 426, 168, 64, 24, 9),
  type = "P"
)

poissonNormalFit = fit(
  poissonNormalData,
  model = externalPoissonNormalModel(),
  nterms = 9
)

c(mu = poissonNormalFit$mu, sigma = poissonNormalFit$sigma)
```

The same common fitted-object and model-comparison APIs remain available even though every probability evaluation requires downstream numerical integration.

```{r poisson-normal-generic-apis}
fitted(poissonNormalFit)
predict(poissonNormalFit, newdata = 0:8, interval = "none")
logLik(poissonNormalFit)
deviance(poissonNormalFit)
AIC(poissonNormalFit)
BIC(poissonNormalFit)
```

## Serialization and retained model objects

New `psFit` objects retain their originating model descriptor. That avoids reconstructing third-party models from a hard-coded list of model names.

```{r retained-model}
class(poissonFit$modelObject)
class(poissonNormalFit$modelObject)
```

This also means a fitted object can survive ordinary R serialization, provided the corresponding S3 methods are available when the object is used again.

```{r serialization}
path = tempfile(fileext = ".rds")
saveRDS(poissonNormalFit, path)
restoredFit = readRDS(path)
logLik(restoredFit)
```

A downstream package therefore needs to remain installed and loadable when its fitted objects are restored, just as with other S3-based extension systems.

## What an extension package should contain

A real external package does not need to copy fitPS fitting code. Its distribution layer can remain small:

1. a constructor returning `psModel(...)` with a unique subclass;
2. S3 methods for the required public fitPS generics;
3. any private helpers used to evaluate probabilities, likelihoods, integrals, or parameter transformations;
4. `S3method(...)` entries in its `NAMESPACE`; and
5. behavioural tests showing that `fit()`, fitted probabilities, prediction, and relevant model-comparison criteria work.

For an MLE-only model, there is no requirement to implement a posterior engine. Bayesian support should be declared only when the model supplies the corresponding engine-specific contract.

## Design lessons from the two examples

The Poisson example shows that a distribution with a closed-form probability mass function can be added without another public fitting function and without any fitPS source modification.

The Poisson-normal example demonstrates something stronger. Its marginal probability and likelihood require numerical integration, yet that integration, its starting-value logic, and all distribution-specific mathematics remain external. `fitPS` still provides optimizer orchestration and the standard `psFit` interface. No Poisson-normal branch is needed in `fit()`, `predict()`, `logLik()`, AIC, BIC, or deviance.

That is the intended extension boundary: **new statistical models supply model mathematics; fitPS supplies common fitting and fitted-object infrastructure.**
