---
title: "Fitting a Zeta distribution to a P survey&mdash;number of groups data"
output: rmarkdown::html_vignette
bibliography: '`r system.file("REFERENCES.bib", package="fitPS")`'
vignette: >
  %\VignetteIndexEntry{Fitting a Zeta distribution to a P survey&mdash;number of groups data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
suppressWarnings(suppressPackageStartupMessages({library(fitPS)}))
```

In this example we will learn how to use `fitPS` to fit a Zeta distribution to some data from a survey where the number of groups of glass found is recorded. The data in this example comes from @roux2001 who surveyed the footwear of 776 individuals in south-eastern Australia, and is summarised in the table below.
```{r results='asis', echo=FALSE}
knitr::kable(Psurveys$roux$data, 
             col.names = c("$n$", "$r_n$"),
             align = "rr")
```
This data set is built into the package and can be accessed from the `Psurveys` object. That is, we can type:
```{r}
roux = Psurveys$roux
```
The 

